---
title: "course-project-1"
author: "Michał Sawicki"
date: "2023-08-25"
output: html_document
---
``` {r message=FALSE}
library(dplyr)
library(scales)
Sys.setlocale("LC_TIME", "en_US.UTF-8")
```
## **Loading and preprocessing the data**

### Read the data...
```{r}
data_activities <- read.csv("activity.csv")
head(data_activities)
```
## **What is mean total number of steps taken per day?**

### Calculate the total number of steps taken per day
```{r}
steps_per_day <- data_activities %>% group_by(date) %>% summarize(sum(steps, na.rm=TRUE))
colnames(steps_per_day) <- c("day", "number_of_steps")
head(steps_per_day)
```

### Histogram of the total number of steps taken each day
```{r}
hist(steps_per_day$number_of_steps, col="steelblue", xlab="Number of Steps", main="")
```

### Mean and median of the total number of steps taken per day
```{r}
mean(steps_per_day$number_of_steps)
median(steps_per_day$number_of_steps)
```

## **What is the average daily activity pattern?**

### Calculate the average number of steps taken from 5 minutes intervals
```{r}
steps_per_interval <- data_activities %>% group_by(interval) %>% summarize(mean(steps, na.rm=TRUE))
colnames(steps_per_interval) <- c("time_interval", "average_number_of_steps")
head(steps_per_interval)
```

### Time series plot of the 5-minute interval and the average number of steps taken
```{r}
plot(steps_per_interval$time_interval, steps_per_interval$average_number_of_steps, type="l", xlab="Minute of a Day", ylab="Average Number of Steps")
```

#### According this data... day is more than 1440 minutes long.

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max_steps <- max(steps_per_interval$average_number_of_steps)
filter(steps_per_interval, average_number_of_steps == max_steps)
```

## **Imputing missing values**

### The total number of missing values in the dataset
```{r}
steps <- data_activities$steps
nums_of_missing <- sum(is.na(steps))
total <- length(data_activities$steps)
nums_of_missing
```

### Fill in the missing data with the average value and create a new data set
```{r}
data_activities$steps[is.na(data_activities$steps)] <- mean(data_activities$steps, na.rm=TRUE)
head(data_activities)
```

###  A histogram of the total number of steps taken each day
```{r}
steps_per_day <- data_activities %>% group_by(date) %>% summarize(sum(steps, na.rm=TRUE))
colnames(steps_per_day) <- c("day", "number_of_steps")
hist(steps_per_day$number_of_steps, col="steelblue", xlab="Number of Steps", main="")
```

### Mean and median of the total number of steps taken per day
```{r}
mean(steps_per_day$number_of_steps)
median(steps_per_day$number_of_steps)
```

#### Completing the missing data with the average value has a significant impact on the data parameters because the percentage of values is high and is about:
```{r}
percent(nums_of_missing/total)
```

## **Are there differences in activity patterns between weekdays and weekends?**
```{r}
data_activities$date <- weekdays(as.Date(data_activities$date))
colnames(data_activities) <- c("steps", "day_of_week", "interval")

data_activities <- data_activities %>%
  mutate(day_type = case_when(
    day_of_week %in% c("Monday", "Thuesday", "Wednesday", "Thuesday", "Friday") ~ "weekday",
    day_of_week %in% c("Saturday", "Sunday") ~ "weekend"
  ))
head(data_activities)
```

```{r}
data_weekday <- filter(data_activities, day_type =="weekday")
data_weekend <- filter(data_activities, day_type == "weekend")

steps_per_interval_weekday <- data_weekday %>% group_by(interval) %>% summarize(mean(steps, na.rm=TRUE))
colnames(steps_per_interval_weekday) <- c("time_interval", "average_number_of_steps")
steps_per_interval_weekend <- data_weekend %>% group_by(interval) %>% summarize(mean(steps, na.rm=TRUE))
colnames(steps_per_interval_weekend) <- c("time_interval", "average_number_of_steps")
head(steps_per_interval_weekday)
```

## Plot of average step activities (during the weekdays and weekends)
```{r}
plot(steps_per_interval_weekday$time_interval, steps_per_interval_weekday$average_number_of_steps, type="l", xlab="Minute of a Day", ylab="Average Number of Steps", col="blue")
lines(steps_per_interval_weekend$time_interval, steps_per_interval_weekend$average_number_of_steps, type="l", xlab="Minute of a Day", ylab="Average Number of Steps", col="red")
legend("topright", legend = c("Weekdays", "Weekend"), col = c("blue", "red"), lty = 1, cex = 0.8)
```

#### CONCLUSION: On weekends, activity starts later and is more evenly distributed throughout the day.